---
title: "Turning nCoV case reports into infection incidence"
author: James Hay <jhay@hsph.harvard.edu>
output:
  rmarkdown::github_document:
    pandoc_args: --webtex
---

```{r setup, include=TRUE,message=FALSE,warning=FALSE}
## Assumed parameters
mean_incubation <- 10
var_incubation <- 12.33
repeats <- 1000
latest_date <- as.Date("2020-01-27",origin="1970-01-01")

source("code/analysis.R")
knitr::opts_chunk$set(echo = TRUE) 
```

## Updates
10am 27/01/2019 
* Added same plot facetted by province
* Ran sensitivity with mean incubation period of 5 days


## Introduction
Available line list data for 2019-nCoV mostly show the date of report and not the date of infection onset. These reported cases therefore do not directly represent underlying transmission dynamics, but rather the day at which sick individuals become known to the healthcare system (plus some delay in reporting). To understand the dynamics of the outbreak, it is more informative to use the incidence curve for infections rather than case reports. Here, I try to generate realistic infection and symptom onset incidence curves that makes use of the confirmed case data from Hubei and elsewhere in China. I use a bootstrapping technique to propagate uncertainty through the predictions.

This does *not* provide any new data nor does it provide any sort of forward projections. This is simply to generate a range of plausible infection and symptom onset curves that could have given rise to the reported case data.

Note that I have just seen this [bioarxiv preprint](https://www.biorxiv.org/content/10.1101/2020.01.25.919787v1), so have not included it yet^1^.

## Old method
The expected time of actual infection is given by:

$\text{expected day of infection} = \text{day your case was confirmed} - \text{expected delay from symptoms to confirmation} - \text{expected incubation period}$

The number of cases reported on a given day is therefore the sum of infections that occured on all preceding days multiplied by the probability that those cases entered the hospital (and were therefore reported) on that day of their infection.

Another way of thinking about it is that the number of infections on a given day $t$ is the sum of all cases that happened on all future days $i$, multipled by the probability of a case that started on day $t$ entering healthcare $i-t$ days after infection.

$X(t) = \sum_{i=t}^{t+t_{max}} C(i)*p(i-t)$

where $X(t)$ is the number of infections on day $t$; $C(i)$ is the number of cases reported on day $i$; and $p(i-t)$ is the probability of entering the hospital $i-t$ days since the start of infection

## New method
From the perspective of an individual, we can simulate random infection onset times from the date of their case confirmation. We can do this in two parts:

1. Use the distribution of delays between symptom onset and case confirmation for known cases to generate a distribution of confirmation delays.
2. Assume some form of the incubation period to generate a distribution of times between infection and symptom onset.

For each confirmation date, we can simulate both parts of the delay to get a predicted infection onset time. We can then do this many times to get a distribution of infection onset times for each individual that takes into account uncertainty in the confirmation delay and incubation period distributions.

## Data
Data for the 41 confirmed cases pre- 01/01/2020 have already been deaggregated into symptom onset times, so take a look at these^2^. 

Here, I use all of the line list data compiled at^3^: [https://github.com/beoutbreakprepared/nCoV2019(https://github.com/beoutbreakprepared/nCoV2019)

I use only the data from China (outside and inside Hubei). I do not use data from other countries, as the reporting systems are likely very different.

```{r}
print(head(other_dat_china1))
```

## Parameters
I assumed that the incubation period for 2019-nCoV has a mean of 10 days based on this tweet: [https://twitter.com/SunKaiyuan/status/1221477620501577732](https://twitter.com/SunKaiyuan/status/1221477620501577732)^4^. I assume a variance of 12.33 days based on estimates for SARS^5^.

## Results
### Confirmation delay distribution
First, I calculated the distribution of times between symptom onset and date of confirmation *where known*. I assumed that cases had a reporting delay of *at least* one day (so the couple of dcases with a delay less than this were set to 1 day) Then, I fit a geometric distribution through these data.
```{r}
p_other_confirm_fit
```
The red line shows the geometric distribution fit. Dashed vertical line indicates first day post symptoms that confirmation could occur.

### Hospitalisation delay distribution
Not used for the augmentation step, but for prosperity here's the hospitalisation delay distribution:
```{r}
p_other_hosp_fit
```

### Incubation period distribution
Next, I generated a gamma distribution for the incubation period based on a mean of 10 days and variance of 12.33 days matching previous estimates for SARS^5^.
```{r}
p_incubation
```

### Example augmented dataset
For each confirmed case date, I simulated a delay from time of symptom onset from the first geometric distribution and subtracted this from the confirmation date. Then, for each time of symptom onset (if known or augmented), I simulated a delay for time since infection and subtracted this from the date of symptom onset.
```{r}
p_data_augmented_example
```

### Bootstrapped incidence curves
To propagate the randomness from these distributions and thereby reflect uncertainty in onset times, I repeated this process 1000 times to generate 95\% quantiles on the incidence curves. Note that the infection incidence curve drops towards the end. This is because a large number of infections that occured in the past have not yet been included in the case confirmation counts.
```{r, fig.width=12, fig.height=8}
results_panel_10day <- results_panel
results_panel_10day
```

## Spatial variation 
Same analysis, but with augmented times stratified by province. Note that this reveals that the augmented times use both confirmed cases and those with reported symptom onsets, which is why the orange/blue lines can overtake the grey.
```{r, fig.width=12, fig.height=12}
by_province
```


## Shorter incubation period
```{r, fig.width=12, fig.height=8}
## Assumed parameters
mean_incubation <- 5
var_incubation <- 12.33
latest_date <- as.Date("2020-01-27",origin="1970-01-01")

source("code/analysis.R")
results_panel
```

## Comments
* I have included 100 bootstrapped infection time and symptom onset time data sets in the git repo. These should match the line list data, so can be merged.
* Modifying case data in this way might be useful as sensitivity analyses for models using the case reports as incidence (which is problematic if done naively)
* The reporting delay distribution has definitely changed over time. Again, this could be a function of $t$ rather than a single uniform distribution.

## Acknowledgements
Thanks Amy Dighe and Charlie Whittaker for checking that the logic here makes sense.

## References
1. Liu et al. Transmission dynamics of 2019 novel coronavirus (2019-nCoV). bioRxiv 2020 doi:https://doi.org/10.1101/2020.01.25.919787
2. Huang et al. Clinical features of patients infected with 2019 novel coronavirus in Wuhan, China. Lancet 2020 doi:https://doi.org/10.1016/S0140-6736(20)30183-5
3. https://github.com/beoutbreakprepared/nCoV2019(https://github.com/beoutbreakprepared/nCoV2019
4. https://twitter.com/SunKaiyuan/status/1221477620501577732
5. Cai et al. Refined estimate of the incubation period of severe acute respiratory syndrome and related influencing factors. Am J Epidemiol 2006;163(3):211-6 doi:https://doi.org/10.1093/aje/kwj034

