---
title: "WTF"
output:
  html_document:
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,fig.width=12, fig.height=8)
```

## Set up some fake data
```{r}
library(tidyverse)
library(ggplot2)
library(patchwork)
prob_detect <- 0.15
tmax <- 100
straggly_start <- 0
dates <- 1:(tmax+straggly_start)
cases <- c(rpois(straggly_start, 2), rpois(tmax,floor((1:tmax)^1.8)))
true_onsets <- tibble(date=dates,n=cases,var="true_onsets")
#true_onsets <- true_onsets %>% mutate(n = ifelse(date == 8,n, 0))

generate_delay <- function(tmax, p) {  
  ts <- 0:(tmax-1)
  tmp <- dgeom(ts,p)
  tmp <- tmp/sum(tmp)
  cumu_geom <- cumsum(tmp)
  stop <- runif(1)
  index <- 0
  x <- 0
  while(x < stop) {
    index <- index + 1
    x <- cumu_geom[index]
  }
  ts[index]
}
```

## Naive
```{r}
toy_onsets <- true_onsets
toy_onsets <- toy_onsets %>% mutate(n = ifelse(date == 8,n, 0))
toy_linelist <- toy_onsets %>% uncount(n) %>% select(-var)
colnames(toy_linelist)[1] <- "date_onset"
toy_linelist <- toy_linelist %>% mutate(delay=rgeom(n(), prob_detect)+1,
                                date_confirmation=date_onset + delay,
                                delay_back=rgeom(n(), prob_detect)+1,
                                #delay_back=generate_delay(0, prob_detect)+1,
                                date_onset_back=date_confirmation-delay_back)
toy_linelist <- toy_linelist %>% filter(date_confirmation <= (tmax+straggly_start))
toy_linelist_melt <- toy_linelist %>% pivot_longer(cols=c("date_confirmation","date_onset","date_onset_back"), names_to="var",values_to="date")
sim_datp <- toy_linelist_melt %>% group_by(var, date) %>% tally() %>% bind_rows(toy_onsets) %>% 
  ggplot() + 
  geom_point(aes(x=date,y=n,col=var)) +
  geom_line(aes(x=date,y=n,col=var)) +
  #coord_cartesian(xlim=c(-straggly_start, (tmax+straggly_start))) +
  theme_bw()
sim_datp

```


## Maximum likelihood-ish
```{r}
toy_onsets <- true_onsets
toy_onsets <- toy_onsets %>% mutate(n = ifelse(date == 8,n, 0))
toy_linelist <- toy_onsets %>% uncount(n) %>% select(-var)
colnames(toy_linelist)[1] <- "date_onset"
toy_linelist <- toy_linelist %>% mutate(delay=rgeom(n(), prob_detect)+1,
                                date_confirmation=date_onset + delay,
                                #delay_back=rgeom(n(), prob_detect)+1,
                                delay_back=generate_delay(8, prob_detect)+1,
                                date_onset_back=date_confirmation-delay_back)
toy_linelist <- toy_linelist %>% filter(date_confirmation <= (tmax+straggly_start))
toy_linelist_melt <- toy_linelist %>% pivot_longer(cols=c("date_confirmation","date_onset","date_onset_back"), names_to="var",values_to="date")
sim_datp1 <- toy_linelist_melt %>% group_by(var, date) %>% tally() %>% bind_rows(toy_onsets) %>% 
  ggplot() + 
  geom_point(aes(x=date,y=n,col=var)) +
  geom_line(aes(x=date,y=n,col=var)) +
  #coord_cartesian(xlim=c(-straggly_start, (tmax+straggly_start))) +
  theme_bw()
sim_datp1

```

## Run the forward model
ie. generate confirmation dates from onset dates
```{r}
## Forward model
## At least 1 day delay then geometric wait time
linelist <- true_onsets %>% uncount(n) %>% select(-var)
colnames(linelist)[1] <- "date_onset"
linelist <- linelist %>% mutate(delay=rgeom(n(), prob_detect)+1,
                                date_confirmation=date_onset + delay)
linelist <- linelist %>% filter(date_confirmation <= (tmax+straggly_start))
linelist_melt <- linelist %>% pivot_longer(cols=c("date_confirmation","date_onset"), names_to="var",values_to="date")
sim_datp1 <- linelist_melt %>% group_by(var, date) %>% tally() %>% bind_rows(true_onsets) %>% 
  ggplot() + 
  geom_point(aes(x=date,y=n,col=var)) +
  geom_line(aes(x=date,y=n,col=var)) +
  coord_cartesian(xlim=c(-straggly_start, (tmax+straggly_start))) +
  theme_bw()
sim_datp1
  
```

```{r}

p1 <- linelist %>% ggplot() + geom_histogram(aes(x=delay),binwidth=1) + 
  facet_wrap(~date_onset,scales="free_y") + ggtitle("Forward")
p2 <- linelist %>% ggplot() + geom_histogram(aes(x=delay),binwidth=1) + 
  facet_wrap(~date_confirmation,scales="free_y") + ggtitle("Backward")

p1
```
```{r}
p2
```

## Run the backward model
ie. generate onset dates from confirmation dates
```{r}
## Backward model
linelist_backward <- linelist %>% select(date_confirmation)
true_onsets_long <- true_onsets %>% uncount(n)
linelist_chunks <- NULL
chunk_dates <- c(25,50,75,100)
for(i in seq_along(chunk_dates)){
  print(chunk_dates[i])
  linelist_tmp <- linelist_backward %>% filter(date_confirmation <= chunk_dates[i]) #%>% 
    #mutate(delay=rgeom(n(),prob_detect)+1,date_onset=date_confirmation-delay) %>%
  linelist_tmp$delay <- NA
  for(j in 1:nrow(linelist_tmp)) {
    linelist_tmp$delay[j] <- generate_delay(linelist_tmp$date_confirmation[j],prob_detect)+1
  }
  linelist_tmp$date_onset <- linelist_tmp$date_confirmation - linelist_tmp$delay
  linelist_tmp <- linelist_tmp %>% pivot_longer(cols=c("date_confirmation","date_onset"), names_to="var",values_to="date") %>%
  bind_rows(true_onsets_long) %>% filter(date <= chunk_dates[i]) %>% mutate(chunk=paste0("day <= ",chunk_dates[i]))
  linelist_chunks[[i]] <- linelist_tmp
}
print("done")
linelist_backward_chunks <- do.call("bind_rows", linelist_chunks)
linelist_backward_chunks$chunk <- factor(linelist_backward_chunks$chunk, levels=c("day <= 25", "day <= 50","day <= 75", "day <= 100"))
linelist_backward <- linelist_backward_chunks %>% filter(chunk == "day <= 100") %>% select(-chunk)
sim_datp2 <- linelist_backward_chunks %>% group_by(var, date, chunk) %>% tally() %>%
  ggplot() + 
  geom_point(aes(x=date,y=n,col=var)) +
  geom_line(aes(x=date,y=n,col=var)) +
  #coord_cartesian(xlim=c(-straggly_start, tmax+straggly_start)) +
  facet_wrap(~chunk,scales="free") +
  theme_bw()
sim_datp2
  
```


```{r}

p3 <- linelist_backward %>% filter(var == "date_onset") %>% ggplot() + geom_histogram(aes(x=delay),binwidth=1) + 
  facet_wrap(~date,scales="free_y") + ggtitle("Forward")
p4 <- linelist_backward %>% filter(var == "date_confirmation")%>% ggplot() + geom_histogram(aes(x=delay),binwidth=1) + 
  facet_wrap(~date,scales="free_y") + ggtitle("Backward")
p3
```
```{r}
p4
```

Looks like forward and backward distributions make sense in both cases right?


## Now do exactly the same thing for the backward model with the real data
ie. generate onset dates from confirmation dates
```{r}
combined_dat_final <- read_csv("tmp.csv")
dat_ncov <- combined_dat_final$date_confirmation
real_dat <- tibble(date_confirmation=dat_ncov) 
## Backward model
#real_dat$delay <- NA
#for(i in 1:nrow(real_dat)) {
#  real_dat$delay[i] <- generate_delay(cumu_geom,times)
#}
#real_dat$date_onset <- real_dat$date_confirmation - real_dat$delay

real_dat <- real_dat %>% mutate(delay=rgeom(n(), prob_detect)+1,
                                date_onset=date_confirmation-delay)

dat1 <- real_dat %>% pivot_longer(cols=c("date_confirmation","date_onset"), names_to="var",values_to="date") %>% group_by(var,date) %>% tally() %>%
  ggplot() + geom_line(aes(x=date,y=n, col=var))
dat1

```

```{r}
p5 <- real_dat %>% ggplot() + geom_histogram(aes(x=delay),binwidth=1) + 
  facet_wrap(~date_onset,scales="free_y") + ggtitle("Forward - looks weird")
p6 <- real_dat %>% ggplot() + geom_histogram(aes(x=delay),binwidth=1) + 
  facet_wrap(~date_confirmation,scales="free_y") + ggtitle("Backward - looks fine")
p5
```

```{r}
p6
```

Try again with smoothed data
```{r}
real_dat1 <- real_dat %>% group_by(date_confirmation) %>% tally()
smooth_fit <- smooth.spline(real_dat1, spar=0.5)

real_dat1$n <- floor(smooth_fit$y)

real_dat1 <- real_dat1 %>% uncount(n)

## Backward model
#real_dat1$delay <- NA
#for(i in 1:nrow(real_dat1)) {
#  real_dat1$delay[i] <- real_dat1$delay[i] <- generate_delay(real_dat1$date_confirmation[i],prob_detect)+1
#}
#real_dat1$date_onset <- real_dat1$date_confirmation - real_dat1$delay

real_dat1 <- real_dat1 %>% mutate(delay=rgeom(n(),prob_detect)+1,
                                  date_onset=date_confirmation-delay)


smooth_forward <- real_dat1 %>% ggplot() + geom_histogram(aes(x=delay),binwidth=1) + 
  facet_wrap(~date_onset,scales="free_y") + ggtitle("Forward")
smooth_backward <- real_dat1 %>% ggplot() + geom_histogram(aes(x=delay),binwidth=1) + 
  facet_wrap(~date_confirmation,scales="free_y") + ggtitle("Backward")
smooth_forward

```
```{r}
smooth_backward
```

```{r}
dat2 <- real_dat1 %>% pivot_longer(cols=c("date_confirmation","date_onset"), names_to="var",values_to="date") %>% 
  group_by(var,date) %>% tally() %>%
  ggplot() + geom_line(aes(x=date,y=n, col=var))
dat2
```